Turbo fast no compromise gravitational wave parameter estimation

### 2022-11-02 src/tex/ms.tex
Finish FlowMC

While gradient-based samplers have been shown to be superior in terms of
performance when compared to other gradient-free algorithms in many practical
examples\kw{cite}, there are still a number of classes of problems
gradient-based samplers do not solve well. \footnote{To be specific, we are
referring to sampling algorithms that use first order derivative here. Sampling
algorithms that use information of higher order derivative such as manifold-MALA
and Riemannian-HMC can in principle decorrelate local correlation in the target
distribution, however they often have instability issue when they are used on
real-life application, so they are not used often in practice.} For examples,
target distribution that exhibits local correlation are hard to dealt with,
since by construction first order gradient-based algorithms can only handle
global correlation structure. Another example is multi-modality. If there are
multiple modes in the target distribution, an independent chain will likely to
be trapped in one mode and take extremely long time to transverse between the
modes. This means the relative weights between modes will much longer to sample
compared to the shape of each mode.

Moreover, before we can use the sampling chain to estimate the quantity we care,
the sampler often needs to first find the most probable region in the target
space, which is a common process that is referred to "burn-in" in the
literature. This means one would discard a certain amount of data generated from
the beginning and only use the later part to estimate the quantities of
interest. The burn-in phase of a gradient-based sampler is often as long as the
sampling phase, which means a good portion of computation is not necessary
helpful in estimating the target quantities.

Normalizing flow is a neural-network based technique that aims at learning a
mapping between a simple distribution, such as a Gaussian distribution, to a
complex distribution, often given in the form of data samples from a target
distribution. Once the network is trained, one can evaluate the probability
density of the complex distribution and samples from it very efficiently.
The core of normalizing flow is given by
\begin{align}
    p_x(X) = p_z(Z) \prod_i \left| \frac{\partial f}{\partial x_i}\right|,
\end{align}
where $p_x(X)$ is the complex target distribution $p_z(Z)$ is the simple latent
distribution, and $f$ is a series of invertible parameterized transform that
connect the two distributions. For a detail discussion of the algorithm, we
refer the readers to \kw{cite}.

As mentioned in the previous subsection, one of the main issue gradient-based
sampler is it does not explore local correlation and multi-modality well. This
is exactly where the normalizing flow model can help. While the method would
still work when there is only one independent local sampler chain, this method
works much better when we employ a lot of independent chains. By looking across
multiple chains, the normalizing flow learns the global landscape of the target
distribution, hence we can use it as a \textit{proposal} distribution. This
means independent chains can now jump between modes, or within a single mode
that exhibits complex geometry that cannot be handled easily only through the
local sampler, which greatly reduce both the burn-in time and sampling time.

@@

In this work, the use of accelerators provides two independent perks that
tremendously benefits the parameter estimation process. First, using
accelerators allow us to run many independent chains at the same time, which
benefits the training of the normalizing flow. Since we generate the data we use
to train the normalizing flow on-the-fly, the more independent data we can feed
to the training process, the higher chance the normalizing flow can learn a
reasonable representation of the global landscape of the target distribution. If
we only use a small amount of chains, we are limited to the correlated samples
from each chain, so we have to run more sequential steps to get the same amount
of independent data as compared to running more chains, where former option does
not benefit from parallelization but the latter does. In another word, being
able to use many independent chains help the normalizing flow learn the global
landscape faster in WALL time.

Another benefit i

>>>>

% Problem with just HMC or MALA
While gradient-based samplers have been shown to be superior in terms of
performance when compared to other gradient-free algorithms in many practical
examples\kw{cite}, there are still a number of classes of problems
gradient-based samplers do not solve well. \footnote{To be specific, we are
referring to sampling algorithms that use first order derivative here. Sampling
algorithms that use information of higher order derivative such as manifold-MALA
and Riemannian-HMC can in principle decorrelate local correlation in the target
distribution, however they often have instability issue when they are used on
real-life application, so they are not used often in practice.} For examples,
target distribution that exhibits local correlation are hard to dealt with,
since by construction first order gradient-based algorithms can only handle
global correlation structure. Another example is multi-modality. If there are
multiple modes in the target distribution, an independent chain will likely to
be trapped in one mode and take extremely long time to transverse between the
modes. This means the relative weights between modes will much longer to sample
compared to the shape of each mode.

% Long burn in too
Moreover, before we can use the sampling chain to estimate the quantity we care,
the sampler often needs to first find the most probable region in the target
space, which is a common process that is referred to "burn-in" in the
literature. This means one would discard a certain amount of data generated from
the beginning and only use the later part to estimate the quantities of
interest. The burn-in phase of a gradient-based sampler is often as long as the
sampling phase, which means a good portion of computation is not necessary
helpful in estimating the target quantities.

% Crux of normalizing flow
Normalizing flow is a neural-network based technique that aims at learning a
mapping between a simple distribution, such as a Gaussian distribution, to a
complex distribution, often given in the form of data samples from a target
distribution. Once the network is trained, one can evaluate the probability
density of the complex distribution and samples from it very efficiently.
The core of normalizing flow is given by
\begin{align}
    p_x(X) = p_z(Z) \prod_i \left| \frac{\partial f}{\partial x_i}\right|,
\end{align}
where $p_x(X)$ is the complex target distribution $p_z(Z)$ is the simple latent
distribution, and $f$ is a series of invertible parameterized transform that
connect the two distributions. For a detail discussion of the algorithm, we
refer the readers to \kw{cite}.

% How does it help our sampling
As mentioned in the previous subsection, one of the main issue gradient-based
sampler is it does not explore local correlation and multi-modality well. This
is exactly where the normalizing flow model can help. While the method would
still work when there is only one independent local sampler chain, this method
works much better when we employ a lot of independent chains. By looking across
multiple chains, the normalizing flow learns the global landscape of the target
distribution, hence we can use it as a \textit{proposal} distribution. This
means independent chains can now jump between modes, or within a single mode
that exhibits complex geometry that cannot be handled easily only through the
local sampler, which greatly reduce both the burn-in time and sampling time.

@@

% More sample helps training normalizing flow
In this work, the use of accelerators provides two independent perks that
tremendously benefits the parameter estimation process. First, using
accelerators allow us to run many independent chains at the same time, which
benefits the training of the normalizing flow. Since we generate the data we use
to train the normalizing flow on-the-fly, the more independent data we can feed
to the training process, the higher chance the normalizing flow can learn a
reasonable representation of the global landscape of the target distribution. If
we only use a small amount of chains, we are limited to the correlated samples
from each chain, so we have to run more sequential steps to get the same amount
of independent data as compared to running more chains, where former option does
not benefit from parallelization but the latter does. In another word, being
able to use many independent chains help the normalizing flow learn the global
landscape faster in WALL time.

% GPU helps pack a shit ton of waveform evaluation
Another benefit accelerator brings is parallel evaluation of waveforms.
The most expensive part of evaluating the likelihood is generating the waveform.
This is especially significant when analysis long signals such as BNS merger,
since the cost of evaluating the waveform scales linearly with the number of time samples.
While the most accurate waveform model, numerical relativity, is a sequential process,
most of the gravitational wave models used in PE are approximants, meaning they are fits to numerical relativity waveforms.
These fitted form can be evaluated at any given time or frequency without the need to simulate the signals from an initial condition.
This means generating a waveform from an approximant can in fact be benefited from parallelization.
Together with heterodyned likelihood, we can evaluate the likelihood at $\mathcal{O}(10^7)$ different locations on an Nvidia A100 GPU.
The high throughput of likelihood evaluations unlocks the potential of the \texttt{flowMC} sampling algorithms.
